A mechanization of semi-simplicial and semi-cubical sets in the calculus of inductive constructions

### 2022-12-17 paper/paper.tex
paper: spell-check document

\begin{example}[$\DeltaPlus^2$]
  In dimension $2$, the standard augmented semi-simplex is drawn as two points, given by $\Hom(1, 2)$, along with a line connecting them, given by $\Hom(2, 2)$. We use the color black to denote the unique morphims in $\Hom(0, 1)$ and $\Hom(0, 2)$.

@@

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a : A, (B a \rightarrow C)$ and $(\Sigma a : A, B a) \rightarrow C$. Let us illustrate how we build $\fullframe[n]$ by refining types of $X_0, X_1, \ldots, X_{n - 1}$ so that they have the form $\fullframe[n] \rightarrow \U$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Each $\fullframe[n]$ is made of $n$-layers, written $\layer[n][p]$ with $p < n$, initalized with the $\unittype$ type. Then, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$. The rest of the construction follows.

@@

Notably, the definition of $\restrl$ relies on a definitional equality expressing the commutation of the composition of $\restrf$. This is apriori undecidable, and we have to prove the commutation propositionally before using the reflection rule. Proving this itself requires an induction on the dimension, and on the structure of $\framep$, $\layer$, and $\filler$. This is what $\cohf$ proves, as shown in the table \ref{tab:coh}, using auxiliary definitions $\cohl$ and $\cohc$. Even though it looks independent of the other tables, it has to proved mutually with the definitions of $\framep$, $\layer$, $\filler$, and their corresponding restrictions.

@@

There are a few more subtleties. In $\cohf$ and $\cohc$, we use the reflection rule to identify equalities of pairs with a conjunction of two equalities. In $\cohl$, we use the reflection rule to identify equality of functions with pointwise equality. We additionally rely heavily on basic properties of the groupoid transport, which we don't detail.

@@

Remark: If we weren't working in \U, but in \HGpd, we would need to prove one more higher-dimensional coherence, and if we were working in \Type, we would need to prove arbitarily many higher-dimensional coherences. For the unfamiliar reader, $\HGpd$ is the subset of types $A$ such that for all $x$ and $y$ in $A$, $x =_{\mathit{HGpd}} y$ is in \U. See \cite{herbelin15} for a discussion on the need for recursive higher-dimensional coherence conditions in formulating higher-dimensional structures in type theory.

@@

Since the construction shown in the previous sections are by induction on $n$, and dependencies are on lower $n$ and $p < n$, one would imagine formalizing this using well-founded induction in the calculus of inductive constructions. We initially tried this approach: we had terms dependent on the proofs of the case distinction of $n' \leq n$ implies $n' < n$ or $n' = n$, and these proofs were didn't have definitional computational rule; it was necessary to reason proposionally on the computational property of the case distinction, and it even turned out to be unmanageable. Hence, we chose a different route: in practice, since $\restrf[n]$ depends on $\framep[n]$ and $\framep[n-1]$, while $\cohf[n]$ depends on $\framep[n]$, $\framep[n-1]$, and $\framep[n-2]$, we only need to keep track of three ``levels'', and we built separate data structures for the levels, with dependencies. More concretely, we build the ten definitions shown in the tables by induction, and this is part of the definition of a larger record. Other fields in the record are necessary because we need copies of $\framep$, $\layer$, $\filler$ at levels $n - 1$ and $n - 2$, $\restrf$, $\restrl$, and $\restrc$ at level $n - 1$, and equations to recall the definitions of these objects at lower levels.

The entire construction relies on inequalities over natural numbers, and we have three different definitions of $\leq$ addressing different concerns, and conversions betwen them. First, we have what we call the ``recusive variant'', phrased as:

@@

\texttt{leY} is an improvement over \texttt{leR} since reflexivity is now definitionally the neutral elemement of transitivity, and associativity of transitivity also holds definitionally. Although it significantly eases our proof, there are some instances where unification is unable to solve for the existentials, and we have to provide them explicitly.

>>>>

\begin{example}[$\DeltaPlus^2$]
  In dimension $2$, the standard augmented semi-simplex is drawn as two points, given by $\Hom(1, 2)$, along with a line connecting them, given by $\Hom(2, 2)$. We use the color black to denote the unique morphisms in $\Hom(0, 1)$ and $\Hom(0, 2)$.

@@

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a : A, (B a \rightarrow C)$ and $(\Sigma a : A, B a) \rightarrow C$. Let us illustrate how we build $\fullframe[n]$ by refining types of $X_0, X_1, \ldots, X_{n - 1}$ so that they have the form $\fullframe[n] \rightarrow \U$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Each $\fullframe[n]$ is made of $n$-layers, written $\layer[n][p]$ with $p < n$, initialized with the $\unittype$ type. Then, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$. The rest of the construction follows.

@@

Notably, the definition of $\restrl$ relies on a definitional equality expressing the commutation of the composition of $\restrf$. This is a priori undecidable, and we have to prove the commutation propositionally before using the reflection rule. Proving this itself requires an induction on the dimension, and on the structure of $\framep$, $\layer$, and $\filler$. This is what $\cohf$ proves, as shown in the table \ref{tab:coh}, using auxiliary definitions $\cohl$ and $\cohc$. Even though it looks independent of the other tables, it has to proved mutually with the definitions of $\framep$, $\layer$, $\filler$, and their corresponding restrictions.

@@

There are a few more subtleties. In $\cohf$ and $\cohc$, we use the reflection rule to identify equalities of pairs with a conjunction of two equalities. In $\cohl$, we use the reflection rule to identify equality of functions with point-wise equality. We additionally rely heavily on basic properties of the groupoid transport, which we don't detail.

@@

Remark: If we weren't working in \U, but in \HGpd, we would need to prove one more higher-dimensional coherence, and if we were working in \Type, we would need to prove arbitrarily many higher-dimensional coherences. For the unfamiliar reader, $\HGpd$ is the subset of types $A$ such that for all $x$ and $y$ in $A$, $x =_{\mathit{HGpd}} y$ is in \U. See \cite{herbelin15} for a discussion on the need for recursive higher-dimensional coherence conditions in formulating higher-dimensional structures in type theory.

@@

Since the construction shown in the previous sections are by induction on $n$, and dependencies are on lower $n$ and $p < n$, one would imagine formalizing this using well-founded induction in the calculus of inductive constructions. We initially tried this approach: we had terms dependent on the proofs of the case distinction of $n' \leq n$ implies $n' < n$ or $n' = n$, and these proofs were didn't have definitional computational rule; it was necessary to reason propositionally on the computational property of the case distinction, and it even turned out to be unmanageable. Hence, we chose a different route: in practice, since $\restrf[n]$ depends on $\framep[n]$ and $\framep[n-1]$, while $\cohf[n]$ depends on $\framep[n]$, $\framep[n-1]$, and $\framep[n-2]$, we only need to keep track of three ``levels'', and we built separate data structures for the levels, with dependencies. More concretely, we build the ten definitions shown in the tables by induction, and this is part of the definition of a larger record. Other fields in the record are necessary because we need copies of $\framep$, $\layer$, $\filler$ at levels $n - 1$ and $n - 2$, $\restrf$, $\restrl$, and $\restrc$ at level $n - 1$, and equations to recall the definitions of these objects at lower levels.

The entire construction relies on inequalities over natural numbers, and we have three different definitions of $\leq$ addressing different concerns, and conversions between them. First, we have what we call the ``recursive variant'', phrased as:

@@

\texttt{leY} is an improvement over \texttt{leR} since reflexivity is now definitionally the neutral element of transitivity, and associativity of transitivity also holds definitionally. Although it significantly eases our proof, there are some instances where unification is unable to solve for the existentials, and we have to provide them explicitly.
